---
title: "Project"
author: "Amir Ragaie - Ebrahim AlaaEldin"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}

if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
if (!requireNamespace("SNPRelate", quietly=TRUE))
    BiocManager::install("SNPRelate")
if (!requireNamespace("GENESIS", quietly=TRUE))
    BiocManager::install("GENESIS")
if (!requireNamespace("SeqVarTools", quietly=TRUE))
    BiocManager::install("SeqVarTools")
if (!requireNamespace("gdsfmt", quietly=TRUE))
    BiocManager::install("gdsfmt")
if (!requireNamespace("GWASTools", quietly=TRUE))
    BiocManager::install("GWASTools")

if (!requireNamespace("openxlsx", quietly=TRUE))
    install.packages("openxlsx")

if (!requireNamespace("qqman", quietly=TRUE))
    install.packages("qqman")

library(SNPRelate)
library(GENESIS)
library(SeqVarTools)
library(gdsfmt)
library(GWASTools)
library(openxlsx)
library(qqman)



```

## Table of Contents

1.  [Task 1: Compute Kinship using SNPRelate and GENESIS]
2.  [Task 2: Compute mQTLs with Mixed Models]
3.  [Task 3: Inflation factor calculation]
4.  [Task 4: Manhattan Plot]
5.  [Task 5: Metabolic Networks]
6.  [Task 6: Annotate Significant SNPs]
7.  [Task 7: Regional plots using SNIPA]

\newpage
## Task 1: Compute Kinship using SNPRelate and GENESIS
a. Compute kinship using the SNPRelate and GENESIS package in R. You can use the
IBD family of functions, but note that you need to transform the plink format to GDS
format using the function snpgdsPED2GDS that takes the ped and map files.


```{r}
# convert to GDS format
snpgdsPED2GDS(ped.fn = "Qataris 156 Dataset\\Qatari156.ped",
              map.fn = "Qataris 156 Dataset\\Qatari156.map",
              out.gdsfn = "Qatari_SNPs.gds")

```

```{r}
# Open the GDS file
genofile <- snpgdsOpen("Qatari_SNPs.gds")

# Calculate kinship using MLE method (recommended for accuracy)
ibd_mle <- snpgdsIBDMLE(genofile, num.thread = 16, kinship = TRUE)

# Extract kinship matrix and pairs dataframe
kinship_matrix <- ibd_mle$kinship
ibd_df <- snpgdsIBDSelection(ibd_mle)

# Find all pairs where kinship > 0.1, excluding diagonal
pairs_idx <- which(kinship_matrix > 0.1, arr.ind = TRUE)
pairs_idx <- pairs_idx[pairs_idx[, 1] != pairs_idx[, 2], ]

# To avoid double counting pairs (i,j) and (j,i), divide count by 2
num_pairs <- nrow(pairs_idx) / 2

cat("Number of pairs with kinship > 0.1:", num_pairs, "\n")

# Get unique individuals involved in these pairs
unique_individuals <- unique(c(pairs_idx[,1], pairs_idx[,2]))
cat("Number of unique individuals with kinship > 0.1:", length(unique_individuals), "\n")

# Close the GDS file after kinship calculation
snpgdsClose(genofile)

```



# GDS (Genomic Data Structure)

**GDS** stands for **Genomic Data Structure**. It is a file format used to efficiently store and access large-scale genetic data, such as SNP genotype data.



| Method                                                                                                   | Best for                                          |
|----------------------------------------------------------------------------------------------------------|---------------------------------------------------|
| [snpgdsIBDKING](https://www.rdocumentation.org/packages/SNPRelate/versions/1.6.4/topics/snpgdsIBDKING)   | Quick, robust kinship estimates on large datasets |
| [snpgdsIBDMLE](https://www.rdocumentation.org/packages/SNPRelate/versions/1.6.4/topics/snpgdsIBDMLE)     | More accurate, maximum likelihood kinship         |

**Input:** You give it an IBD result object from `snpgdsIBDMLE`, which contains IBD/kinship estimates between individuals.

**Output:** It returns a data frame (`ibd_df`) listing pairs of individuals with their IBD estimates and related information.
b. Report the number of individuals who have a kinship > 0.1

\newpage
## Task 2: Compute mQTLs with Mixed Models
a. Compute mQTLs using all SNPs and all metabolites to identify associations and report
their p-values. You will need to use mixed models to include kinship.
```{r}
# Load PCA results (adjust paths accordingly)
eigenvectors <- read.table("Qataris 156 Dataset\\pca_results.eigenvec", header = FALSE)
eigenvalues <- read.table("Qataris 156 Dataset\\pca_results.eigenval", header = FALSE)

colnames(eigenvectors) <- c("FID", "IID", paste0("PC", 1:(ncol(eigenvectors)-2)))
pcs <- eigenvectors[, 2:5]

# Load metabolites phenotype data (adjust path)
metabolites_matrix <- read.csv("qatari_metabolites_2025.csv", stringsAsFactors = FALSE)

```

## Prepare data
```{r}
# Merge phenotype with PCs by sample IID (assumed matching sample IDs)
seqData <- merge(metabolites_matrix, pcs, by.x = "Sample", by.y = "IID")
seqData$scanID <- seqData$Sample

# Create ScanAnnotationDataFrame object
scanAnnot <- ScanAnnotationDataFrame(seqData)

# Subset kinship matrix for samples - make sure kinship_matrix is loaded from Task 1
kin <- kinship_matrix[seqData$scanID, seqData$scanID]


```


b. GENESIS provides a fitNullModel function that computes residuals of metabolites from
the covariates, including kinship, and an assocTestSingle function that computes the
mQTLs.

## Fit null model for each metabolite
```{r}
results_list <- list()
heritability_list <- data.frame(Metabolite = character(), Heritability = numeric(), stringsAsFactors = FALSE)

for (metab in colnames(metabolites_matrix)[-1]) {  # Assuming first column is Sample ID
  phenotype <- metabolites_matrix[[metab]]
  
  # Fit null model with kinship and PCs as covariates
  nullmod <- fitNullModel(
    x = scanAnnot,
    outcome = phenotype,
    covars = c("PC1", "PC2", "PC3"),
    cov.mat = kin,
    family = "gaussian"
  )
  
  # Run association tests (SNP by SNP)
  assoc <- assocTestSingle(nullmod, scanAnnot) # or seqData, depending on format
  
  # Filter significant associations with p < 1e-4
  sig_assoc <- assoc[assoc$Score.pval < 1e-4, ]
  if (nrow(sig_assoc) > 0) {
    sig_assoc$Metabolite <- metab
    results_list[[metab]] <- sig_assoc
  }
  
  # Estimate heritability and save
  hci <- varCompCI(nullmod)
  heritability_list <- rbind(heritability_list, 
                             data.frame(Metabolite = metab, Heritability = hci[1, "Proportion"]))
}


```

c. Include the first three principal components in the covariates.
d. Report significant SNP-Metabolite associations until p < 0.0001 (that’s three zeros after
the decimal point) in an Excel sheet, showing beta values, standard error, effect allele,
degree of freedom, and p values. (all values are the direct output of assocTestSingle).
e. Report heritability from the varCompCI function in GENESIS.
```{r}
library(openxlsx)

# Combine all significant associations into one dataframe
final_sig_assoc <- do.call(rbind, results_list)

# Select and rename columns to match requirements
report_df <- final_sig_assoc[, c("Metabolite", "variant.id", "beta", "se", "effect.allele", "df", "Score.pval")]
colnames(report_df) <- c("Metabolite", "SNP", "Beta", "SE", "Effect_Allele", "DF", "P_value")

write.xlsx(list(
  Significant_mQTLs = report_df,
  Heritability = heritability_list
), file = "mQTL_Results.xlsx", rowNames = FALSE)

```



\newpage
## Task 3: Inflation factor calculation

```{r}
inflation_factors <- data.frame(Metabolite = character(), Lambda = numeric(), stringsAsFactors = FALSE)

for (metab in names(results_list)) {
  assoc <- results_list[[metab]]
  
  chisq <- qchisq(1 - assoc$Score.pval, df = 1)
  lambda <- median(chisq, na.rm = TRUE) / qchisq(0.5, df = 1)
  
  inflation_factors <- rbind(inflation_factors, data.frame(Metabolite = metab, Lambda = lambda))
}

# Print average inflation factor
cat("Average inflation factor (lambda):", mean(inflation_factors$Lambda, na.rm = TRUE), "\n")

```

\newpage
## Task 4: Manhattan Plot
```{r}
# Combine all association results (significant or not)
all_assoc <- do.call(rbind, results_list)

manhattan_data <- data.frame(
  SNP = all_assoc$variant.id,
  CHR = as.numeric(as.character(all_assoc$chr)),
  BP = all_assoc$pos,
  P = all_assoc$Score.pval,
  stringsAsFactors = FALSE
)

# Clean and filter data
manhattan_data <- na.omit(manhattan_data)
manhattan_data <- manhattan_data[!is.na(manhattan_data$CHR), ]

# Plot Manhattan plot with significance line at p = 1e-4
manhattan(
  manhattan_data,
  genomewideline = -log10(1e-4),
  suggestiveline = FALSE,
  col = c("blue", "orange"),
  main = "Manhattan Plot of mQTL Associations"
)

```

\newpage
## Task 5: Metabolic Networks
\newpage
## Task 6: Annotate Significant SNPs
\newpage
## Task 7: Regional plots using SNIPA
\newpage